Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev

### 2025-11-12 main.tex
Expanded the note on the unstable model at 6 km/h.
Addressed the review comment about how much the cyclist experienced the range of forces and why we didn't try to eliminate the learning effects.

For the 6~\si{\kph} experiment, scaling the proportional feedback gain linearly
with respect to speed stabilizes the normally unstable weave mode of the
bicycle down to 3.4~\si{\kph} for the riderless bicycle and 8.0~\si{\kph} for
the ridden bicycle. For the 10~\si{\kph} experiment, it stabilizes the normally
unstable weave mode of the bicycle down to 3.8~\si{\kph} for the riderless
bicycle and 9.7~\si{\kph} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}.  Notably, the bicycle is not stable at
6~\si{\kph} and just stable at 10~\si{\kph}. Figure~\ref{fig:eigenvalues} also
shows the identified weave mode captured using the methods of
Kooijman~et~al.~\citep{Kooijman2009a} and gives an indication of the
performance of the system, with the model predicting the weave mode well the
region of the tested speeds.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row), controlled with \(\kappa=3.9\) (middle
    row), and controlled with \(\kappa=5.2\) (lower row) root locus of the
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the linear Carvallo-Whipple bicycle model without the inertial effect
    of a rigid rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at:
    6~\si{\kph} and 10~\si{\kph}. The grey shaded region is the linear stable
    speed range. Black stars are experimentally identified weave mode decay and
    frequency.}
  \label{fig:eigenvalues}
\end{figure}

@@

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure applying perturbations
above and below the initial perturbation threshold, while allowing small
progression of the perturbation threshold to accommodate learning effects. Five
possible perturbation forces are determined based on the initial perturbation
threshold estimate: the initial estimate itself, two perturbations lower than
the initial estimate, and two perturbations higher than the initial estimate.
The five possible perturbations are separated by 10~\si{\newton} steps. For
example, if the initial estimate of the perturbation threshold of a participant
is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
100~\si{\newton}. Which one of these five perturbations is chosen first is
determined at random. If the perturbation results in a fall, the estimate of
the perturbation threshold is decreased by 10~\si{\newton}, and vice versa if
the perturbation did not result in a fall. Five new possible perturbation
forces are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied. The goal of this adaptive staircase procedure is to
have participants fall for approximately 50\% of the time. After the first set
of 20 perturbations, we let the cyclist rest and then perform another 20
perturbations. We randomize whether the balance assist system is on or off
during the first or second set of 20 perturbations for each participant and the
participants were not told the order.

>>>>

For the 6~\si{\kph} experiment, scaling the proportional feedback gain linearly
with respect to speed stabilizes the normally unstable weave mode of the
bicycle down to 3.4~\si{\kph} for the riderless bicycle and 8.0~\si{\kph} for
the ridden bicycle. For the 10~\si{\kph} experiment, it stabilizes the normally
unstable weave mode of the bicycle down to 3.8~\si{\kph} for the riderless
bicycle and 9.7~\si{\kph} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}. Notably, the bicycle is not stable at 6~\si{\kph}
but the time constant of the unstable eigenvalue is increased by a factor of 8
and at 10~\si{\kph} the system is just stable. Figure~\ref{fig:eigenvalues}
also shows the identified weave mode captured using the methods of
Kooijman~et~al.~\citep{Kooijman2009a} and gives an indication of the
performance of the system, with the model predicting the weave mode well the
region of the tested speeds.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row), controlled with \(\kappa=3.9\) (middle
    row), and controlled with \(\kappa=5.2\) (lower row) root locus of the
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the linear Carvallo-Whipple bicycle model without the inertial effect
    of a rigid rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at:
    6~\si{\kph} and 10~\si{\kph}. The grey shaded region is the linear stable
    speed range. Black stars are experimentally identified weave mode decay and
    frequency.}
  \label{fig:eigenvalues}
\end{figure}

@@

Following the initial threshold determination where the participant experienced
a full range of pertubation magnitudes, we then choose perturbation forces
according to a random and adaptive staircase procedure applying perturbations
above and below the initial perturbation threshold, while allowing small
progression of the perturbation threshold to accommodate learning effects. We
did not try to elimate learning effects due to the time per participant that
would be required, but factored this into the later analysis as an alternative.
Five possible perturbation forces are determined based on the initial
perturbation threshold estimate: the initial estimate itself, two perturbations
lower than the initial estimate, and two perturbations higher than the initial
estimate.  The five possible perturbations are separated by 10~\si{\newton}
steps. For example, if the initial estimate of the perturbation threshold of a
participant is 80~\si{\newton}, the five possible perturbations are 60, 70, 80,
90 and 100~\si{\newton}. Which one of these five perturbations is chosen first
is determined at random. If the perturbation results in a fall, the estimate of
the perturbation threshold is decreased by 10~\si{\newton}, and vice versa if
the perturbation did not result in a fall. Five new possible perturbation
forces are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied. The goal of this adaptive staircase procedure is to
have participants fall for approximately 50\% of the time. After the first set
of 20 perturbations, we let the cyclist rest and then perform another 20
perturbations. We randomize whether the balance assist system is on or off
during the first or second set of 20 perturbations for each participant and the
participants were not told the order.
